# Paired tests {#sec-pair}


```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
source("R/booktem_setup.R")
source("R/my_setup.R")

```

### Packages

```{r}
library(tidyverse)
library(GGally)
library(emmeans)
library(performance)
library(broom.helpers)
```

```{r, eval=TRUE, echo=FALSE}

library(tidyverse)

library(here)

darwin <- read_csv(here("files", "darwin.csv"))

```


The structure of our linear model so far has produced the output for a standard two-sample Student's *t*-test. However, when we *first* [calculated our estimates by hand](#differences-between-groups) - we started by making an average of the paired differences in height. To generate the equivalent of a paired *t*-test, we simply have to add the factor for pairs to our linear model formula:

::: {.panel-tabset}

## Base T

```{r}
lsmodel_darwin <- lm(height ~ type + factor(pair), data = darwin)
summary(lsmodel_darwin)
```

## Tidyverse

```{r}
darwin |>  
  mutate(pair = as_factor(pair)) |>  
  lm(height ~ type + pair, data = _) |>  
  broom::tidy()
```

:::

> Note that I have made pair a factor - pair 2 is not greater than pair 1 - so it does not make sense to treat these as number values.


The table of coefficients suddenly looks a lot more complicated! This is because **now** the intercept is the height of the crossed plant from pair 1:

* The second row now compares the average heights difference of Crossed and Selfed plants **when they are in the same pair** 

* rows three to 16 compare the average difference of each pair (Crossed and Selfed combined) against pair 1

Again the linear model computes every possible combination of *t*-statistic and *P*-value, however the only one we care about is the difference in Cross and Self-pollinated plant heights. If we ignore the pair comparisons the second row gives us a *paired t*-test. 'What is the difference in height between Cross and Self-pollinated plants when we hold pairs constant.'

Q. By including pairs we are making sure we deal with `r mcq(c("Missing data bias", answer = "Omitted variable bias", "Survivorship bias"))`

## Activity 1: Compare paired model designs

We previously compared the difference in plant heights by subtracting the difference between each pair of plants. Check our reasoning on the model design above by making another linear model based on the difference between pairs and check whether this is significantly different from 0, `lm(difference ~ 1)`

`r hide("Solution")`

```{r}

darwin_wide <- darwin |>  
  # pivot/reshape the data from long to wide format
  pivot_wider(names_from = type, values_from = height) |>  
  # Calculate the difference between 'Cross' and 'Self' heights for each pair
  mutate(difference = Cross - Self)

wide_model <- lm(difference ~ 1, data = darwin_wide)

summary(wide_model)
```

`r unhide()`

For completeness let's generate the confidence intervals for the *paired t*-test and compare them to our *unpaired t*-test. 

::: {.panel-tabset}

## Unpaired

```{r}

lm(height ~ type, data = darwin) |> 
   broom::tidy(conf.int=T) |> 
    slice(1:2) 
```

## Paired

```{r}
lm(height ~ type + factor(pair), data = darwin) |>  
  broom::tidy(conf.int=T) |> 
  slice(1:2) # just show first two rows
```

:::

We can see that estimate of the mean difference is identical but the 95% confidence intervals are now slightly different. So in this particular version we have actually increased our level of uncertainty by including the pair parameter. 

```{r}

m1 <- lm(height ~ type, data = darwin) |>  
  broom::tidy( conf.int=T) |>  
  slice(2:2) |>  
  mutate(model="unpaired")

m2 <- lm(height ~ type + factor(pair), data = darwin) |>  
  broom::tidy(conf.int=T) |>  
  slice(2:2) |>  
  mutate(model="paired")

rbind(m1,m2) |>  
  ggplot(aes(model, estimate))+
  geom_pointrange(aes(ymin=conf.high, ymax=conf.low))+
  geom_hline(aes(yintercept=0), linetype="dashed")+
  theme_minimal()+
  coord_flip()

```

```{block, type = "info"}
Choosing the right model

In this case we had a good *a priori* reason to include pair in our model, so I would argue that it should stay - alternative approaches are model simplification and stepwise removal. We will discuss these later. 
```

## Effect sizes

We have discussed the importance of using confidence intervals to talk about effect sizes. When our 95% confidence intervals do not overlap the intercept, this indicates we have difference in our means which is significant at $\alpha$ = 0.05. More interestingly than this it allows us to talk about the 'amount of difference' between our treatments, the lower margin of our confidence intervals is the smallest/minimum effect size. On the response scale of our variables this is very useful, we can report for example that there is *at least* a 0.4 inch height difference between self and crossed fertilised plants at $\alpha$ = 0.05. 


## Activity 2: Write-up


Q. Can you write a summary of the **Results**?



`r hide("Solution")`

We analysed the difference in heights between selfed and cross-pollinated maize plants using a paired analysis of 15 pairs of plants. On average we found plants that have been cross pollinated (20.2 inches [18.3, 22], (mean [95% CI])) were taller than the self-pollinated plants (t(14) = -2.148, p = 0.0497), with a mean difference in height of 2.62 [0.004, 5.23] inches .

`r unhide()`


## Summary

This chapter finally allowed us to calculate *P*-values and test statistical significance for our experiments using linear models. We also compared the linear model structures for producing a paired vs. unpaired *t*-test. 

A single experiment is never definitive, and a reliance on reporting *P*-values is uninformative and can be misleading. Instead reporting estimates and confidence intervals allows us to report our levels of uncertainty, and provides results which are more informative for comparative studies. 


