# Factors {#sec-factors}


```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
source("R/booktem_setup.R")
source("R/my_setup.R")
penguins_clean_names <- readRDS("data/penguin_clean_names.RDS")
```

:::{.callout-important collapse = "true"}

## Penguin clean names dataset

```{r, eval = FALSE}
penguins_clean_names <- readRDS(url("https://github.com/Philip-Leftwich/Oct-Intro-Analytics/raw/refs/heads/dev/data/penguin_clean_names.RDS"))

```

:::


In R, factors are a class of data that allow for **ordered categories** with a fixed set of acceptable values. 

Typically, you would convert a column from character or numeric class to a factor if you want to set an intrinsic order to the values (“levels”) so they can be displayed non-alphabetically in plots and tables, or for use in linear model analyses (more on this later). 

```{r}

levels(penguins_clean_names$species)

```


Working with factors is easy with the `forcats` package:

```{r, eval = F}
penguin_factors <- penguins_clean_names |> 
  mutate(species = as_factor(species))

levels(penguin_factors$species)

```


### Applying changes across columns

Using `across` - we can apply functions to columns based on selected criteria - here within mutate we are changing each column in the `.cols` argument and applying the function `forcats::as_factor()`

```{r}

penguins_clean_names |> 
  mutate(
    across(.cols = c("species", "region", "island", "stage", "sex"),
           .fns = forcats::as_factor)
  ) |> 
  select(where(is.factor)) |> 
  glimpse()

```

::: {.callout-important}

Unless we assign the output of this code to an R object it will just print into the console, in the above I am demonstrating how to change variables to factors but we aren't "saving" this change.

:::


### Ordering factors manually

With the `forcats::fct_relevel` function we can set factors and apply a specified level at the same time:

```{r}
penguins_clean_names |> 
  mutate(species = fct_relevel(species, 
                               "Adelie Penguin (Pygoscelis adeliae)", 
                               "Chinstrap penguin (Pygoscelis antarctica)", 
                               "Gentoo penguin (Pygoscelis papua)")) |> 
  ggplot(aes(x = species))+
  geom_bar()+
  coord_flip()
  

```

### Ordering factors by another variable

With the function `forcats::fct_infreq` we can change the order according to how frequently each level occurs

```{r}
penguins_clean_names |> 
   mutate(species = fct_infreq(species)) |>  
  ggplot(aes(x = species))+
  geom_bar()+
  coord_flip()
  

```

The `forcats::fct_rev()` function in R is used to reverse the order of levels in a factor variable. It is particularly useful for changing the order of factor levels when you want to display data in a reversed or descending order.

```{r}
penguins_clean_names |> 
  mutate(species = fct_rev(as_factor(species))) |> 
  ggplot(aes(x = species))+
  geom_bar()+
  coord_flip()

```

```{r, warning = FALSE}
penguins_clean_names |> 
  mutate(species = as_factor(species) |> 
           fct_reorder(body_mass_g,
                       .fun = mean,
                       .na_rm = T)) |> 
  # by default the levels are ordered by the median values of the continuous variable
  # mean, min and max can all be included here
  ggplot(aes(x = species,
             y = body_mass_g,
             colour = species))+
  geom_boxplot(width = .2,
               outlier.shape = NA)+
  coord_flip()+
  theme(legend.position ="none")
```

### Factor bins

If we want to we can also *bin* continuous data into useful chunks: 

```{r, eval = T, warning = F, message = F}
penguins_clean_names <- penguins_clean_names |> 
  mutate(mass_range = case_when(
    body_mass_g <= 3500 ~ "smol penguin",
    body_mass_g >3500 & body_mass_g < 4500 ~ "mid penguin",
    body_mass_g >= 4500 ~ "chonk penguin",
    .default = NA)
  )
```

If we make a barplot, the order of the values on the x axis will typically be in alphabetical order for any character data

```{r, eval = T, warning = F, message = F}
penguins_clean_names |> 
  drop_na(mass_range) |> 
  ggplot(aes(x = mass_range))+
  geom_bar()

```


:::{.task}
::::{.task-header}
Your turn
::::
::::{.task-container}

* [ ] Our new variable `mass_range` is not yet set as a factor, can you try this now?

`r hide("Solution")`

To convert a character or numeric column to class factor, you can use any function from the `forcats` package. They will convert to class factor and then also perform or allow certain ordering of the levels - for example using `forcats::fct_relevel()` lets you manually specify the level order. 

The function `as_factor()` simply converts the class without any further capabilities.

```{r, eval = T, warning=FALSE, message = F}
penguins_clean_names <- penguins_clean_names |> 
  mutate(mass_range = as_factor(mass_range))

```


```{r}
levels(penguins_clean_names$mass_range)
```

` r unhide()`


* [ ] Make sure the factors are set to a sensible order

` r hide("Solution")`

Below we use `mutate()` and `as_factor()` to convert the column flipper_range from class character to class factor. 

```{r}
# Correct the code in your script with this version
penguins_clean_names <- penguins_clean_names |> 
  mutate(mass_range = fct_relevel(mass_range, 
                                  "smol penguin", 
                                  "mid penguin", 
                                  "chonk penguin")
         )

levels(penguins_clean_names$mass_range)
```

Now when we call a plot, we can see that the x axis categories match the intrinsic order we have specified with our factor levels. 

```{r, eval = T, warning = F, message = F}
penguins_clean_names |> 
  drop_na(mass_range) |>  
  ggplot(aes(x = mass_range))+
  geom_bar()

```

`r unhide()`

::::
:::

